Competitive differential chemotaxis can explain the self-organisation of retinotopic maps

### 2023-09-15 paper/main.tex
Completes update of initial Results section in which model is described

%% Any aliases here
\newcommand{\mb}[1]{\mathbf{#1}} % this won't work?
% Emphasis and bold.
\newcommand{\e}{\emph}
\newcommand{\code}[1]{\textsf{#1}}
\newcommand{\dvrg}{\nabla\vcdot\nabla}
%% END aliases

@@

\author[1*]{Sebastian~S.~James}
\author[1]{Stuart~P.~Wilson}
\affil[1]{Department of Psychology, The University of Sheffield, Sheffield, United Kingdom.}
\corr{seb.james@sheffield.ac.uk}{SSJ}
%% END OF PREAMBLE

@@

A key test of the validity of recent models is their ability to account for data from a set of genetic manipulation studies by \cite{brown_topographic_2000} and \cite{reber_relative_2004}, in which axons from a subset of retinal ganglia cells treated with a Eph-A3 knock-in form a separate, rostrally shifted topographic map, which becomes indistinct from that formed by untreated cells, at a ‘collapse point’ on the rostral side. \cite{Sterratt2013} found that a mixture of counter-gradients and an additional competition term provided the best account of these data. A follow-up study in which a range of different computational models were systematically compared \citep{hjorth_quantitative_2015} found that one focused on competition
% Stuart: I think the model is Koulakov's, not Triplett's. In Triplett2011 they give evidence for a competition model (like Koulakov's) by reducing the number of RGCs in the Math5 mutant mouse. *Hjorth* shows that the Koulakov model is the best one at reproducing the collapse in the EphA3 ki experiment.
by \cite{Triplett2011} could best account for several such experimental findings, and that the other types of model could not explain the collapse point.

The model of \cite{Triplett2011} describes maps as minima of an energy function, whose structure can be obtained via a stochastic optimisation process. Accordingly, the energy is the sum of three terms, relating to the influences of chemical gradients, axon density (competition), and activity-dependent growth processes. Maps that optimise this function can be obtained by starting with a random map and then substituting connections at random, accepting those substitutions with a probability that increases if they reduce the energy, and repeating until the algorithm converges on a pattern of connections corresponding to a minimal energy state.

@@

We adopt the same form for the expression of receptors on retinal ganglion cells as
\citet{simpson_simple_2011}:
\begin{equation} \label{e:f}
  f(x,y) = 1.05 + 0.26 \exp(2.3 u),
\end{equation}
where $u$ gives the direction with which the expression increases and may be substituted by $(1-x)$, $(1-y)$, $x$ or $y$. We defined $x$ as an axis from temporal to the nasal retina and $y$ from ventral to dorsal retina.

%% * Simpson and Goodhill modelled according to deltax = B + C + G (Eq 2)
In \citet{simpson_simple_2011}'s model of chemotaxis and competition, the position, $\mathbf{x}_t$, of each branch $b$ on a simulated tectum can be updated according to $\mathbf{x}_{t+1} = \mathbf{x}_{t} + \Delta \mathbf{x}$ with the movement vector given by
% Maybe re-order the terms?
\begin{equation} \label{e:dX}
 \Delta \mathbf{x} = \mathbf{B} + \mathbf{C} + \mathbf{G}.
\end{equation}
$\mathbf{B}$ is a movement applied close to the tissue boundary that ensures all branches remain within the tissue (\citet{holt_target_1998}---see method details).
%
%% * They had C composed of two terms, distance based and a component based on EphA expression
%
$\mathbf{C}$ is the movement due to competitive interactions with other branches
%
and $\mathbf{G}$ is the movement vector given rise to by the chemotaxis effect.

@@

Simpson \& Goodhill's model had a $\mathbf{C}$ comprised of two components; a simple distance-based competition between \emph{all} branches, along with an axon-axon interaction based on the relative levels of the receptor EphA expressed by each interacting axon.

%Although the model incorporated EphA expression, the distance-based competitive interaction was given no mechanism of operation.

To provide an explicit mechanism of operation for the competition, we modelled it as a sum of interactions that are `switched on' via receptor-ligand signalling, meaning that some axons would have no competitive interaction, while others would repel strongly.
This single-component competitive movement vector for a branch $b$ is

\begin{equation} \label{e:X}
\mathbf{C} = \frac{m_{\!_X}}{|B_{b}|} \sum_k^{nM} \hat{\mathbf{x}}_{kb}\,W\,Q(d_{kb}, \mathbf{r}_{b}, \mathbf{l}_{k}).
\end{equation}
%
$m_{\!_X}$ is a scalar parameter, $nM$ is the total number of branches and $\hat{\mathbf{x}}_{kb}$ is the unit vector from branch $k$ to $b$.
%
$W$ is the distance based weighting ($W = 1-\frac{d_{kb}}{2r_{\!_X}}~\mathrm{if}~  d_{kb}\leq 2r_{\!_X}$, otherwise $0$) and $B_{b}$ is the set of branches that are within interaction distance ($2 r_{\!_X}$) of $b$. $r_{\!_X}$ is the receptor-ligand interaction radius.
%
The distance-based weighting can be interpreted as being related to the probability with which a ligand may bind to a receptor and mediate the interaction.
%
$Q$ is a signalling threshold function which depends on the distance $d_{kb}$ between branches $b$ and $k$, the receptor expression on $b$ ($\mathbf{r}_b$) and the ligand expression on $k$ ($\mathbf{l}_k$):
%
\begin{equation}
Q(d_{kb}, \mathbf{r}_{b}, \mathbf{l}_{k}) = \begin{cases}
                 0 & \mathrm{if}~-F_i\,r_{i,b}\,l_{i,k} <
                 s,\,\forall{i}~\mathrm{or}~d_{kb} > 2r_{\!_X} \\
                 1 & \mathrm{otherwise.}
     \end{cases}
\end{equation}
%
$l_{i,k}$ (the expression of ligand $i$ on branch $k$) is the $i^{\mathrm{th}}$ element of $\mathbf{l}_k$ and $r_{i,b}$ is the $i^{\mathrm{th}}$ element of $\mathbf{r}_b$.
$F_i$, defined in Eq.\,\ref{e:G}, is -1 if the $r_{i}\,l_{i}$ interaction signals repulsion and +1 if it signals attraction.
$s$ is a positive signal threshold for interaction.
With the constraint $s>0$ comes the assumption that attractive interactions will have no effect on the competitive movement vector---a receptor-ligand pair that signals attraction cannot cause $Q$ to take the value 1.
$Q$ is set to 1 only if at least one repulsive signal exceeds $s$ and the distance from branch $b$ to branch $k$ is smaller than $2 r_{\!_X}$.

@@

* The S\&G chemotaxis was a placeholder because they chose G to be a vector
  pointing to a predetermined target location. In order to make this a local
  model we do the following....

We also imagine ligand expression on the tectum (L) - describe these here.

@@

%%%%% Describing the simulation TAKE THIS TO METHOD DETAILS
We simulated the retina and tectum as unit square regions and defined retinal receptor expression and tectal ligand expression across these squares in the directions defined in Fig.\,\ref{f:ex}.
The simulated retina possessed $n$=400 (20$\times$20) retinal ganglion cells; the tectum was similarly arranged as 20$\times$20 discrete, square elements, with defined levels of expression for each ligand.
One spatial unit corresponds to 2~mm on the retina or tectum for mouse \citep{reber_relative_2004}.
%%%%%%%%%%%%%

% Keep, but say M grown cones. (M=4 specified in method details)
% ALSO: Put M near competition equation where it's used!
Each retinal ganglion cell (RGC) projects $M=4$ growth cones (referred to as \emph{branches}) which carry a set of $N$ receptors, $r_i$, indexed by $i$ and expressed at levels determined by the cell soma's location on the retinal grid.

The expression of each receptor varies only with respect to one spatial dimension.

% Thi sis in the righ tplace here
We defined 4 receptor expression gradients arranged in orthogonal pairs with the gradient of $r_0$ being orthogonal to that of $r_1$. $r_2$, whose gradient is opposite to $r_0$, was orthogonal to $r_3$ (Fig.\,\ref{f:ex}B). Thus the model is of \emph{differential chemotaxis}.

% NO NEED FOR This sentence/bit
Because there is convincing evidence that EphA and EphB receptors are expressed in exponentially increasing patterns \citep{reber_relative_2004,feldheim_genetic_2000,brown_topographic_2000,koulakov_stochastic_2004}, we use an exponential form for retinal receptor expressions, in common with other modelling studies~\citep{reber_relative_2004,koulakov_stochastic_2004,simpson_simple_2011}.
%
We adopt the same precise form for the retinal receptor expression as \citet{simpson_simple_2011}, $f(x,y) = 1.05 + 0.26 \exp(2.3 u)$, where $u$ gives the direction with which the expression increases and is substituted by $(1-x)$ for $r_0$, $(1-y)$ for $r_1$, $x$ for $r_2$ and by $y$ for $r_3$ (Fig.\,\ref{f:ex}).

@@

While we do not explicitly name $r_0$, $r_1$, etc.~as \emph{EphA}, \emph{EphB}, the suggestion is that $r$ could include EphA, EphB, Ryk \citep{schmitt_wntryk_2006} and Neogenin \citep{rajagopalan_neogenin_2004} receptors and that $L$ could include the ephrin-A, ephrin-B, Wnt3 \citep{schmitt_wntryk_2006} and RGM \citep{monnier_rgm_2002} ligands, each of which has been shown to play a r\^ole in retinotectal map formation.
%%%%%%%%%%%%%%%%%

\subsection*{Simulation} % MOST OF THIS GOES TO METHOD DETAILS, just keep
                         % first sentence, have list of values used in sims, then 'see method details',

The model has four free parameters; $m_{\!_G}$, $m_{\!_X}$, $r_{\!_X}$ and $s$. We fixed $m_{\!_G}$ to the value $0.0012$ and used a numerical optimization process to find the competition parameters that would balance this level of chemotaxis.
We performed a simulated annealing optimization to minimize an objective function for the wildtype pattern (Fig.\,\ref{f:GJ}E) at simulation time 1000, reasoning that evolution would optimize the arrangement of this phenotype.

@@

@@

@@

>>>>

%% Any aliases here
\newcommand{\mb}[1]{\mathbf{#1}} % this won't work?
% Emphasis and bold.
\newcommand{\e}{\emph}
\newcommand{\code}[1]{\textsf{#1}}
\newcommand{\dvrg}{\nabla\vcdot\nabla}

\definecolor{colcmnt}        {rgb} {0.7098, 0.0745, 0.0431}
\newcommand{\redcmnt}[1]{\textcolor{colcmnt}{#1}}

%% END aliases

@@

\author[1]{Sebastian~S.~James}
\author[1*]{Stuart~P.~Wilson}
\affil[1]{Department of Psychology, The University of Sheffield, Sheffield, United Kingdom.}
\corr{s.wilson@sheffield.ac.uk}{SW}
%% END OF PREAMBLE

@@

A key test of the validity of recent models is their ability to account for data from a set of genetic manipulation studies by \cite{brown_topographic_2000} and \cite{reber_relative_2004}, in which axons from a subset of retinal ganglia cells treated with a Eph-A3 knock-in form a separate, rostrally shifted topographic map, which becomes indistinct from that formed by untreated cells, at a ‘collapse point’ on the rostral side. \cite{Sterratt2013} found that a mixture of counter-gradients and an additional competition term provided the best account of these data. A follow-up study in which a range of different computational models were systematically compared \citep{hjorth_quantitative_2015} found that \redcmnt{one focused on competition
% Stuart: I think the model is Koulakov's, not Triplett's. In Triplett2011 they give evidence for a competition model (like Koulakov's) by reducing the number of RGCs in the Math5 mutant mouse. *Hjorth* shows that the Koulakov model is the best one at reproducing the collapse in the EphA3 ki experiment.
by \cite{Triplett2011}} could best account for several such experimental findings, and that the other types of model could not explain the collapse point.

\redcmnt{The model of \cite{Triplett2011} describes maps as minima of an energy function}, whose structure can be obtained via a stochastic optimisation process. Accordingly, the energy is the sum of three terms, relating to the influences of chemical gradients, axon density (competition), and activity-dependent growth processes. Maps that optimise this function can be obtained by starting with a random map and then substituting connections at random, accepting those substitutions with a probability that increases if they reduce the energy, and repeating until the algorithm converges on a pattern of connections corresponding to a minimal energy state.

@@

Each of $n$ retinal ganglion cells (RGC) projects $M$ growth cones (referred to as \emph{branches}) which carry a set of $N$ different receptor types, expressed at levels determined by the cell soma's location on the retinal grid. RGC growth cones and cells on the tectum are also assumed to express $N$ different ligand types which interact with RGC receptors.
%
We adopt the same form for the expression of receptors by retinal ganglion cells as
\citet{simpson_simple_2011}:
\begin{equation} \label{e:f}
  f(x,y) = 1.05 + 0.26 \exp(2.3 u),
\end{equation}
where $u$ gives the direction with which the expression increases and may be substituted by $(1-x)$, $(1-y)$, $x$ or $y$. We defined $x$ as an axis from temporal to the nasal retina and $y$ from ventral to dorsal retina.

%% * Simpson and Goodhill modelled according to deltax = B + C + G (Eq 2)
In Simpson \& GoodHill's model of chemotaxis and competition, the position, $\mathbf{x}_t$, of each branch $b$ on a simulated tectum can be updated according to $\mathbf{x}_{t+1} = \mathbf{x}_{t} + \Delta \mathbf{x}$ with the movement vector given by
% Maybe re-order the terms?
\begin{equation} \label{e:dX}
 \Delta \mathbf{x} = \mathbf{B} + \mathbf{C} + \mathbf{G}.
\end{equation}
$\mathbf{B}$ is a movement applied close to the tissue boundary that ensures all branches remain within the tissue (\citet{holt_target_1998}---see method details).
%
%% * They had C composed of two terms, distance based and a component based on EphA expression
%
$\mathbf{C}$ is the movement due to competitive interactions with other branches
%
and $\mathbf{G}$ is the movement vector given rise to by the chemotaxis effect.

@@

Simpson \& Goodhill's model had a $\mathbf{C}$ comprised of two components; a simple distance-based competition between \emph{all} branches, along with an axon-axon interaction based on the relative levels of the receptor EphA expressed by each interacting axon.
%Although the model incorporated EphA expression, the distance-based competitive interaction was given no mechanism of operation.
To provide an explicit mechanism of operation for the competition, we modelled it as a sum of interactions that are `switched on' via receptor-ligand signalling, meaning that some axons would have no competitive interaction, while others would repel strongly.
This single-component competitive movement vector for a branch $b$ is

\begin{equation} \label{e:X}
\mathbf{C} = \frac{m_{\!_X}}{|B_{b}|} \sum_k^{nM} \hat{\mathbf{x}}_{kb}\,W\,Q(d_{kb}, \mathbf{r}_{b}, \mathbf{l}_{k}).
\end{equation}
%
$m_{\!_X}$ is a scalar parameter and $B_{b}$ is the set of branches that are within interaction distance ($2 r_{\!_X}$) of $b$. $r_{\!_X}$ is the receptor-ligand interaction radius. $nM$ is the total number of branches in the system and $\hat{\mathbf{x}}_{kb}$ is the unit vector from branch $k$ to $b$.
%
$W$ is the distance based weighting ($W = 1-\frac{d_{kb}}{2r_{\!_X}}~\mathrm{if}~  d_{kb}\leq 2r_{\!_X}$, otherwise $0$).
%
The distance-based weighting can be interpreted as being related to the probability with which a ligand may bind to a receptor and mediate the interaction.
%
$Q$ is a signalling threshold function which depends on the distance $d_{kb}$ between branches $b$ and $k$, the receptor expression on $b$ ($\mathbf{r}_b$) and the ligand expression on $k$ ($\mathbf{l}_k$).
$Q$ is set to 1 only if at least one repulsive signal exceeds a positive signal threshold, $s$, and the distance from branch $b$ to branch $k$ is smaller than $2 r_{\!_X}$:
%
\begin{equation}
Q(d_{kb}, \mathbf{r}_{b}, \mathbf{l}_{k}) = \begin{cases}
                 0 & \mathrm{if}~-F_i\,r_{i,b}\,l_{i,k} <
                 s,\,\forall{i}~\mathrm{or}~d_{kb} > 2r_{\!_X} \\
                 1 & \mathrm{otherwise.}
     \end{cases}
\end{equation}
%
$l_{i,k}$ (the expression of ligand $i$ on branch $k$) is the $i^{\mathrm{th}}$ element of $\mathbf{l}_k$ and $r_{i,b}$ is the $i^{\mathrm{th}}$ element of $\mathbf{r}_b$.
$F_i$, defined in Eq.\,\ref{e:G}, is -1 if the $r_{i}\,l_{i}$ interaction signals repulsion and +1 if it signals attraction.
With the constraint $s>0$ comes the assumption that attractive interactions will have no effect on the competitive movement vector---a receptor-ligand pair that signals attraction cannot cause $Q$ to take the value 1.

@@

\citet{simpson_simple_2011} chose a placeholder mechanism for the chemotaxis effect, using a global supervisor to set $\mathbf{G}$ to a vector which would always point to the predetermined target location for each branch.
In order to make this a local model we replaced the global supervisor with an interaction between RGC receptors and ligands expressed on the tectum. We used forms for the ligand expression, $L(\mathbf{x})$, which matched those used for the RGC receptor expression.

@@

Each of the $M$ RGC growth cones carry a set of $N$ receptors, $r_i$, indexed by $i$ and expressed at levels determined by the cell soma's location on the retinal grid.
%
%The expression of each receptor varies only with respect to one spatial dimension.
%
We defined 4 receptor expression gradients (each varying only wrt one spatial dimension) arranged in orthogonal pairs with the gradient of $r_0$ being orthogonal to that of $r_1$. $r_2$, whose gradient is opposite to $r_0$, was orthogonal to $r_3$ (Fig.\,\ref{f:ex}B). Thus the model is of \emph{differential chemotaxis}.

% NO NEED FOR This sentence/bit (keep comment in case for discussion)
%Because there is convincing evidence that EphA and EphB receptors are expressed in exponentially increasing patterns \citep{reber_relative_2004,feldheim_genetic_2000,brown_topographic_2000,koulakov_stochastic_2004}, we use an exponential form for retinal receptor expressions, in common with other modelling studies~\citep{reber_relative_2004,koulakov_stochastic_2004,simpson_simple_2011}.
%
% Already said this:
%We adopt the same precise form for the retinal receptor expression as \citet{simpson_simple_2011}, $f(x,y) = 1.05 + 0.26 \exp(2.3 u)$, where $u$ gives the direction with which the expression increases and is substituted by $(1-x)$ for $r_0$, $(1-y)$ for $r_1$, $x$ for $r_2$ and by $y$ for $r_3$ (Fig.\,\ref{f:ex}).

@@

%While we do not explicitly name $r_0$, $r_1$, etc.~as \emph{EphA}, \emph{EphB}, the suggestion is that $r$ could include EphA, EphB, Ryk \citep{schmitt_wntryk_2006} and Neogenin \citep{rajagopalan_neogenin_2004} receptors and that $L$ could include the ephrin-A, ephrin-B, Wnt3 \citep{schmitt_wntryk_2006} and RGM \citep{monnier_rgm_2002} ligands, each of which has been shown to play a r\^ole in retinotectal map formation.
%%%%%%%%%%%%%%%%%

\subsubsection*{Simulation}

The model has four free parameters; $m_{\!_G}$, $m_{\!_X}$, $r_{\!_X}$ and $s$. We fixed $m_{\!_G}$ to the value $0.0012$ and used a numerical optimization process to find the competition parameters that would balance this level of chemotaxis. The parameters found were $m_{\!_X} = 0.2078$, $r_{\!_X} = 0.0759$ and $s = 4.815$ (see method details).

@@

\subsubsection*{Outtakes}

Previously in results:

While we do not explicitly name $r_0$, $r_1$, etc.~as \emph{EphA}, \emph{EphB}, the suggestion is that $r$ could include EphA, EphB, Ryk \citep{schmitt_wntryk_2006} and Neogenin \citep{rajagopalan_neogenin_2004} receptors and that $L$ could include the ephrin-A, ephrin-B, Wnt3 \citep{schmitt_wntryk_2006} and RGM \citep{monnier_rgm_2002} ligands, each of which has been shown to play a r\^ole in retinotectal map formation.

@@

\subsection*{Simulation domain}
We simulated the retina and tectum as unit square regions and defined retinal receptor expression and tectal ligand expression across these squares in the directions defined in Fig.\,\ref{f:ex}.
The simulated retina possessed $n$=400 (20$\times$20) retinal ganglion cells; the tectum was similarly arranged as 20$\times$20 discrete, square elements, with defined levels of expression for each ligand.
One spatial unit corresponds to 2~mm on the retina or tectum for mouse \citep{reber_relative_2004}.

% Make sure this ain't a copy of results
Each retinal ganglion cell (RGC) projected $M=4$ growth cones (referred to as \emph{branches}) which carry a set of $N$ receptors, $r_i$, indexed by $i$ and expressed at levels determined by the cell soma's location on the retinal grid.

The expression of each receptor was varied only with respect to one spatial dimension.

% Thi sis in the right place here
We defined 4 receptor expression gradients arranged in orthogonal pairs with the gradient of $r_0$ being orthogonal to that of $r_1$. $r_2$, whose gradient is opposite to $r_0$, was orthogonal to $r_3$ (Fig.\,\ref{f:ex}B). Thus the model is of \emph{differential chemotaxis}.

@@

\subsection*{Receptor and Ligand expression forms}

A section to add more detail on the functions used, if necessary.

\subsection*{Parameter finding}

We performed a simulated annealing optimization to find three free parameters; $m_{\!_X}$, $r_{\!_X}$ and $s$ which would balance a chosen evel of chemotaxis strength given by $m_{\!_G}=0.00120$.
We minimized an objective function for the wildtype pattern (Fig.\,\ref{f:GJ}E) at simulation time 1000, reasoning that evolution would optimize the arrangement of this phenotype.
